Handbook streamFind: backend

### 2022-10-17 handbooks/Handbook_backend_streamFind.Rmd
LFU18

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 9, cache = TRUE)
```
<br> <br>

***

@@

[R](https://cran.r-project.org/) and [RTools](https://cran.r-project.org/bin/windows/Rtools/) are essential dependencies. The `r R.version.string` can be obtained in https://cran.r-project.org/. The RTools can be downloaded in https://cran.r-project.org/bin/windows/Rtools/. Make sure to download the right version for the R version installed. Installation instructions are given in both sources.

@@

* [ProteoWizard](https://proteowizard.sourceforge.io/), required for conversion of vendor MS formats to mzML/mzXML;  
* [Java JDK](https://www.oracle.com/java/technologies/downloads/), mandatory for patRoon when plotting structure using MetFrag;  
* [OpenMS](https://www.openms.de/downloads/); esternal tool for peak picking, grouping and alignment;  
* [MetFrag CL](http://ipb-halle.github.io/MetFrag/projects/metfragcl/), used for formula annotation in patRoon;  
* [PubChemLite](https://zenodo.org/record/6503754#.YtqxE3bP1hE), database used with MetFrag in patRoon;  
* [SIRIUS](https://bio.informatik.uni-jena.de/software/sirius/), tool for peak picking and formula annotation in patRoon;    
* [BioTransformer](https://bitbucket.org/djoumbou/biotransformer/downloads/), used for prediction of transformation products in patRoon;  
* [SAFD](https://bitbucket.org/SSamanipour/safd.jl/src/master/), tool for peak picking with data in profile mode;  
* [Open Babel](http://openbabel.org/wiki/Main_Page), used in patRoon for validation of chemicals properties (e.g., InChI and formulae);

@@

The following code lines are adapted from the patRoon [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites) to ensure that both patRoon and other mutual R dependencies are installed properly.

@@

After installation of the patRoon R package, the location of each third party software should be given in an `.Rprofile` file as suggested in section [other dependencies](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#other-dependencies) the patRoon [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites). For facilitation the same code lines are given below. The function `patRoon::verifyDependencies()` from patRoon can be used to verify if all dependencies are found.

@@

The streamFind package can be installed from the GitHub repository https://github.com/ricardobachertdacunha/streamFind with the following code line.

@@

@@

The streamFind package includes example data files via the streamFindData R package. Once installed the full file paths of MS data files can be obtained as shown below. Alternatively, the files can be obtained directly from folder `system.file(package = "streamFindData", dir = "extdata")`. The files available are described in Table \@ref(tab:description-files). Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The centroided Agilent Q-TOF files were trimmed with retention time between 900 and 1350 seconds, _m/z_ for MS1 data between 200 and 450 Da and _m/z_ for MS2 between 35 and 450 Da. Additionally, MS1 and MS2 traces below 100 and 50 counts were removed, respectively. The profile Agilent Q-TOF files were trimmed with retention time between 1000 and 1200 seconds, _m/z_ for MS1 data between 200 and 300 Da and _m/z_ for MS2 between 35 and 300 Da. Additionally, MS1 and MS2 traces below 1 count were removed to exclude empty spectra. The files are used within the document for demonstration of the backend.

@@

```{r show-table-chemicals, echo=FALSE}
db$ionization <- "positive"
db[, `:=`("mz_pos" = neutralMass + 1.0073, "mz_neg" = neutralMass - 1.0073 )]
db[grepl("neg", db$comment), ionization := "both"]
data.table::setnames(db, "mixIUTA", "tag")
db[tag %in% 1, tag := "CS"]
db[, in_file := "1-6"]
db[tag %in% "IS", in_file := "1-24"]

knitr::kable(db[, .(name, CAS, formula, neutralMass, rt, ionization, tag, in_file)], caption = "Chemical standards spiked.") %>%
  kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

The streamFind package uses S4 class objects (e.g., _streamSet_, _msData_, _msAnalysis_, etc.) for structuring data. The S4 classes and their main getter and setter methods are presented and exemplified in this section. An overview of the S4 classes and their hierarchy is presented in Figure \@ref(fig:classes-overview) using MS analyses as an example. Other data types, such as from UV and Raman analyses, would lead to different classes (e.g., _uvData_ and _uvAnalysis_ and _ramanData_ and _ramanAnalysis_, respectively) but the hierarchy is similar as shown for MS data.

@@

* _streamSet_ refers to the broader S4 class applied in the streamFind package and holds basic information for an analysis set, see \@ref(streamset-class);
* _msData_ refers to the S4 subclass of _streamSet_ when MS analyses (i.e., MS files) are added to the set, see \@ref(msdata-class);
* _msAnalysis_ refers to the S4 class of the object generated for each MS file added to the _msData_ object in the slot "analyses";
* `setInfo` is the method to obtain the basic information (i.e., title, date and path) of a _streamSet_ object;
* `analyses` is the name of the slot holding the analyses (as list) and is also a method to get the analysis names in the _streamSet_;
* `replicate` is the name of an analysis replicate group and the method to get the replicate names;
* `blank` is the name of a blank analysis replicate group to be used for blank (blind) subtraction and the method to get blank replicate names;
* __traces__ corresponds to a raw data signal (e.g., mass trace in MS data) obtained from a given analysis file;
* `spectra` for MS analyses is the method to get a table with a spectrum for each cycle time (i.e., retention time);
* __EIC__ is an extracted ion chromatogram (retention time vs intensity) and can be obtained with the method `EICs`;
* __TIC__ is a total ion chromatograms (retention time vs intensity) and can be obtained with the method `TICs`;
* __XIC__ is a three dimensional ion chromatogram (_m/z_ vs retention time vs intensity) and can be obtained with the method `XICs`;
* __peak__ corresponds to a cluster of traces (e.g., integrated chromatographic MS peak) within an analysis file and can be accessed with the method `peaks`; 
* __feature__ corresponds to a group of corresponding peaks across analysis files/analyses and can be accessed with the method `features`;
* __MS2__ corresponds to second order fragmentation data as acquired in tandem MS or MS/MS mode and can be accessed with the method `MS2s`.

@@

The set creation is demonstrated below for the three MS files (triplicate, 1 to 3 in \@ref(tab:description-files)) where a set of chemicals were spiked (see \@ref(spiked-chemicals) with tag IS and S). The set path can be assigned as argument but for the example below it was left as default (i.e., the working directory). The initial S4 class created in the `newStreamSet` function is _streamSet_. The subclass is then defined according to the type of files added. Note that multiple file types are not possible as a subclass cannot be defined.

@@

Because the added files were all with mzML format (i.e., MS files), the subclass is directly defined as _msData_. The structure for further processing data is then set for MS data. Other data formats/structure will lead to different sub classes (e.g., _uvData_ and _ramanData_) but is not yet implemented within the streamFind package.

@@

The _msData_ structure includes the _streamSet_ class plus the `features` slot. The title, date and path slots correspond to the set title, creation date and directory in disk, respectively. The analyses slot is a list with a _msAnalysis_ S4 class object per MS file added. The structure of the _msAnalysis_ class is described in the next sub-chapter. The `features` slot is a _msFeatures_ S4 class object to hold the results from alignment, grouping and annotation of peaks across analyses. Further information for _msAnalysis_ is given in the next sub-chapter (\@ref(msanalysis-class)) and for _msFeatures_ is given in the sub-chapter \@ref(msfeatures-class), after creation of features by performing alignment and grouping of peaks across analyses.

* _msData_ structure:

@@

The _msAnalysis_ class is structured as shown below. Note that the S4 method `getAnalyses` for _msData_ was used to obtain the _msAnalysis_ of the first analysis as defined by the second argument. In general, access to raw data (i.e., traces and chromatograms) as well as respective plotting S4 methods can be applied to both the _msAnalysis_ and _msData_ (as described in \@ref(accessing-raw-data)).

The slot analysis corresponds to the analysis name. The metadata slot is a list of objects with informative entries of the analysis file. The metadata list can be expanded with additional information about the analysis/sample (e.g., sample location, weather conditions, process sensor data, etc.). The slot parameters is used to store processing steps (e.g., peak picking) as _settings_ S4 class objects, including the processing step name/ID, the algorithm used and the respective settings applied. The parameters can be used for history track but also for re-run the processing steps when required. The spectra and chromatograms slots are used to store low level data (i.e., mass traces either in profile or centroided modes) from the MS files. Finally, the peaks slot is used to store the chromatographic peaks after peak picking. As no other processing step was applied the spectra, chromatograms and peaks tables in the _msAnalysis_ are empty for now.

 * _msAnalysis_ structure:

@@

An _msAnalysis_ object can be created without the need of running `newStreamSet`. The method `newAnalysis` can be used for initiating the _msAnalysis_ when a full path of an MS file (i.e., with extension mzML or other MS format) is given. Similar to `newStreamSet`, the format of the file given as file argument dictates the class of the analysis object returned from the `newAnalysis` function.

@@

As the method `getAnalysis` can be used to obtain an _msAnalysis_ from an _msData_ object, the method `addAnalyses` is used to add extra _msAnalysis_ objects to _msData_. Below the files 7 to 9 are added to the already created _msData_ (containing files 1 to 3). Note that both `getAnalyses` and `addAnalyses` can be applied to equivalent classes to _msData_ (i.e., _uvData_ and _ramanData_). Also, note that when adding analyses to an _msData_ object with features (i.e., with peaks grouped across analyses), the _msFeatures_ content is cleared as the grouping is invalidated with new _msAnalysis_ objects.

@@

Below, the `getMetadata` S4 method of _msData_ is used to display the metadata. Note, `getMetadata` also works for _msAnalysis_ S4 class as shown below.

* get metadata:

@@

Additional metadata can be added with the `addMetadata` method for both _msAnalysis_ and _msData_. Note that for _msData_ the metadata input needs to be a list of vectors with the same length as the number of analyses or a data.frame/data.table with the number of rows as the nmumber of analyses, as shown below.

* add metadata to an _msAnalysis_ object (a1):

@@

* add metadata to an _msData_ object (sp1):

@@

The tools for raw data access and plotting (i.e., spectra and chromatograms) share a target search approach. The data search is based on pre-defined time and mass (inc. _m/z_) targets. For traces (i.e., raw MS mass traces), time (in seconds) and _m/z_ (in Da) ranges are given for accessing and plotting. Access to mass traces is relevant for further development of data processing steps/functions and for visualization of raw data, easing the evaluation of processing steps (e.g., peaks wrongly not found during peak picking).

@@

As aforementioned, access to raw data is based on defining _m/z_ and retention time targets. The function `makeTargets` is used for comprehensively build targets. As shown below for Carbamazepin d10 and Diuron d6, multiple ways can be used to assemble targets but similarly returning a uniform `data.table` class object with a target list. Accessing and plotting S4 methods in streamFind use the `makeTargets` function to collect raw data from the analysis files. Therefore, the arguments `mz`, `rt`, `ppm`, `sec` and `id` are present in most S4 methods for accessing and plotting data as demonstrated in the following sections.

```{r carb-diu}
carb <- db[name %in% "Carbamazepin d10", .(name, neutralMass, rt)]
carb$mz <- carb$neutralMass + 1.0073
carb

diu <- db[name %in% "Diuron d6", .(name, neutralMass, rt)]
diu$mz <- diu$neutralMass + 1.0073
diu
```

```{r make_targets}
#case 1
mz1 <- c(247.1650, 239.0620)
rt1 <- c(1075, 1157)
id1 <- c("target1", "target2")
targets1 <- makeTargets(mz = mz1, rt = rt1, ppm = 20, sec = 30, id = id1)

@@

#case 2
mz2 <- data.frame(mzmin = c(247.1601, 239.0572), mzmax = c(247.1699, 239.0668))
rt2 <- data.frame(rtmin = c(1045, 1127), rtmax = c(1105, 1187))
targets2 <- makeTargets(mz = mz2, rt = rt2)

targets2

#case 3
mz3 <- data.frame(
  id = c("target1", "target2"),
  mz = c(247.1650, 239.0620),
  rt = c(1075, 1157)
)
targets3 <- makeTargets(mz = mz3, ppm = 20, sec = 30)

targets3

#case 4
mz4 <- data.frame(
  id = c("target1", "target2"),
  mzmin = c(247.1601, 239.0572), mzmax = c(247.1699, 239.0668),
  rtmin = c(1045, 1127), rtmax = c(1105, 1187)
)
targets4 <- makeTargets(mz = mz4)

targets4
```

@@

Extracted ion chromatograms (EICs) are obtained either from loaded spectra in the _msData_ or _msAnalysis_ objects or from querying traces directly in the raw data file (mzML or mzXML). The latter is less disk memory intense as data is loaded from raw files on demand. The S4 method `EICs` is used to extract MS1 data from raw data files.

@@

The obtained EICs can be plotted with `plotEICs` S4 method, as shown below. The object for `plotEICs` can be a `data.table` as resulting from the `EICs` method or the actual _msData_ or _msAnalysis_ objects, using the same arguments as in `EICs` for collecting the data (i.e., to define the targets).

@@

Total ion chromatograms (TICs) are obtained either by pre-loading and adding the chromatogram data to the _msAnalysis_, extracting the data in a `TICs` call (S4 method) or loading all spectra and applying the sum of intensities to each spectrum. Note, TIC is already available as chromatogram in mzML data from the major HRMS vendors. The S4 method TICs first evaluates if the TIC chromatogram is available in the mzML and when not, the spectra data is extracted and used instead.

`r # TODO add faster functionality for TIC and BPC based on xml2``

@@

A more informative method of obtaining EICs is via the method `XICs` which extract a the three dimensional (_m/Z_, rt and intensity) chromatograms from given targets. In the example below, the _m/z_ dimension is also included in the resulting table to improve inspection of traces.

@@

Similarly, the XICs can be plotted with the S4 method `plotXICs`, as shown below. Other arguments are available for plotting `XICs`, including a target area for the expected targets. As shown in the plot below, the red mark is defined by the arguments `targetsMark` (a table with columns mz and rt to define the expected _m/z_ and retention times) and `ppmMark` and `secMark` to define the mass (in ppm) and time (in seconds) range of the target squares. The `numberRows` is used to control the number of rows of the grouped plot. The concept for the XIC plot is adapted from the R package [MSnbase](https://bioconductor.org/packages/release/bioc/html/MSnbase.html).

@@

The S4 methods applied above parse the data from the raw data files which is less disk memory intense but slower. Alternatively, the raw data can be loaded to (each) _msAnalysis_  and from their is faster available. However, when loaded, the raw data has a higher memory footprint.
Therefore, the data can be optionally loaded and stored as a simple _data.table_ in the respective _msAnalysis_ (other classes of the same level might have the raw data directly loaded if the data has a lower memory footprint). Below we demonstrate how to load (via `loadRawData` S4 method) and access (via `spectra` or `chromatograms`) the raw spectra and chromatograms from an _msAnalaysis_ and _msData_ object. Both S4 methods (`spectra` or `chromatograms`) return a _data.table_ where further operations can be applied to subset the data, as demonstrated below. The standard _data.table_ operations can be applied, see documentation in https://cran.r-project.org/web/packages/data.table/ for more information and code examples/instructions. A cheat sheet for _data.table_ operations can be downloaded in https://raw.githubusercontent.com/rstudio/cheatsheets/main/datatable.pdf. In streamFind, the _data.table_ is always the preferred format for table objects.

```{r loadrawdata-msanalysis}
#load raw data to an msAnalysis object, while setting a minimum intensity value for both MS1 and MS2 traces
a1 <- loadRawData(a1, minIntensityMS1 = 200, minIntensityMS2 = 100)
```

@@

The analysis set creation is demonstrated below for six HRMS data files corresponding to blank and wastewater influent samples, both measured in triplicate. The argument `files` is the main input and can either be a list of full path files, a table with file (i.e., full file path), replicate (i.e., the name of the analysis replicate group for each file) and blank (i.e., the name of the associated blank analysis replicate group for each file), or the `analysisInfo` data.frame from the package [patRoon](https://github.com/rickhelmus/patRoon). The set path can be assign as argument `path` and is the directory where all the set files (i.e., scripts, cache databases, objects and results) will be stored by default.

```{r newstreamset-basicworkflow, message=FALSE, results='hide'}
sp2 <- newStreamSet(files = files[c(13:15, 19:21)], path = getwd(), title = "Project Example")
```

@@

An essential aspect of environmental analysis is to operate with sample/analysis replicates. The replicates in _msData_ can be obtained by the `replicateNames` S4 method as shown below. The setter (assignment) for the replicate names is demonstrated below with the method `replicateNames<-` that takes a vector of character strings with the same length as the number of analyses in a _msData_ object. Alternatively, a table with a replicate character column or a character vector can be given in the files or replicates arguments, respectively. Note that _msAnalysis_ doesn't contain information about replicate correspondence. The correspondence between analyses is storage in the _msFeatures_ within the _msData_.

@@

#setter for replicate names in msData
replicateNames(sp2) <- c(rep("blank", 3),rep("influent", 3))  
replicateNames(sp2)
```

@@

#setter for the blank analysis replicate of msData object
blankReplicateNames(sp2) <- rep("blank", 6)
blankReplicateNames(sp2)
```

@@

`r # TODO list all S4 methods``

@@

#getter for analysis name in msAnalysis
analysisNames(a2)

#getter for polarity of analyses in a msData
#polarities(sp2)

#getter for polarity of an msAnalysis
polarities(a2)

@@

` # TODO create function for collecting MS2 data for peaks/features`

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 9, cache = TRUE)
```

<br> <br>

------------------------------------------------------------------------

@@

[R](https://cran.r-project.org/) and [RTools](https://cran.r-project.org/bin/windows/Rtools/) are essential dependencies. The `r R.version.string` can be obtained in <https://cran.r-project.org/>. The RTools can be downloaded in <https://cran.r-project.org/bin/windows/Rtools/>. Make sure to download the right version for the R version installed. Installation instructions are given in both sources.

@@

-   [ProteoWizard](https://proteowizard.sourceforge.io/), required for conversion of vendor MS formats to mzML/mzXML;\
-   [Java JDK](https://www.oracle.com/java/technologies/downloads/), mandatory for patRoon when plotting structure using MetFrag;\
-   [OpenMS](https://www.openms.de/downloads/); esternal tool for peak picking, grouping and alignment;\
-   [MetFrag CL](http://ipb-halle.github.io/MetFrag/projects/metfragcl/), used for formula annotation in patRoon;\
-   [PubChemLite](https://zenodo.org/record/6503754#.YtqxE3bP1hE), database used with MetFrag in patRoon;\
-   [SIRIUS](https://bio.informatik.uni-jena.de/software/sirius/), tool for peak picking and formula annotation in patRoon;\
-   [BioTransformer](https://bitbucket.org/djoumbou/biotransformer/downloads/), used for prediction of transformation products in patRoon;\
-   [SAFD](https://bitbucket.org/SSamanipour/safd.jl/src/master/), tool for peak picking with data in profile mode;\
-   [Open Babel](http://openbabel.org/wiki/Main_Page), used in patRoon for validation of chemicals properties (e.g., InChI and formulae);

@@

The following code lines are from the patRoon [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites) to ensure that both patRoon and other mutual R dependencies are installed properly.

@@

After installation of the patRoon R package, the location of each third party software should be given in an `.Rprofile` file as suggested in section [other dependencies](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#other-dependencies) of the patRoon [handbook](https://rickhelmus.github.io/patRoon/handbook_bd/manual-installation.html#r-prerequisites). For facilitation the same code lines are given below. The function `patRoon::verifyDependencies()` from patRoon can be used to verify if all dependencies are found.

@@

The streamFind package can be installed from the GitHub repository <https://github.com/ricardobachertdacunha/streamFind> with the following code line.

@@

To be done.

@@

The streamFind package includes example data files via the streamFindData R package. Once installed the full file paths of MS data files can be obtained as shown below. Alternatively, the files can be obtained directly from folder `system.file(package = "streamFindData", dir = "extdata")`. The files available are described in Table \@ref(tab:description-files). Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The centroided Agilent Q-TOF files were trimmed with retention time between 900 and 1350 seconds, *m/z* for MS1 data between 200 and 450 Da and *m/z* for MS2 between 35 and 450 Da. Additionally, MS1 and MS2 traces below 100 and 50 counts were removed, respectively. The profile Agilent Q-TOF files were trimmed with retention time between 1000 and 1200 seconds, *m/z* for MS1 data between 200 and 300 Da and *m/z* for MS2 between 35 and 300 Da. Additionally, MS1 and MS2 traces below 1 count were removed to exclude empty spectra. The files are used within the document for demonstration of the backend.

@@

```{r show-table-chemicals, echo=FALSE}
db$ionization <- "positive"
db[, `:=`("mz_pos" = mass + 1.0073, "mz_neg" = mass - 1.0073 )]
db[grepl("neg", db$comment), ionization := "both"]
data.table::setnames(db, "mix_id", "tag", skip_absent = TRUE)
db[tag %in% 1, tag := "CS"]
db[, in_file := "1-6"]
db[tag %in% "IS", in_file := "1-24"]

knitr::kable(db[, .(name, formula, mass, rt, ionization, tag, in_file)], caption = "Chemical standards spiked.") %>%
  kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

The streamFind package uses S4 class objects (e.g., *streamSet*, *msData*, *msAnalysis*, etc.) for structuring data. The S4 classes and their main getter and setter methods are presented and exemplified in this section. An overview of the S4 classes and their hierarchy is presented in Figure \@ref(fig:classes-overview) using MS analyses as an example. Other data types, such as from UV and Raman analyses, would lead to different classes (e.g., *uvData* and *uvAnalysis* and *ramanData* and *ramanAnalysis*, respectively) but the hierarchy is similar as shown for MS data.

@@

-   *streamSet* refers to the broader S4 class applied in the streamFind package and holds basic information for an analysis set, see \@ref(streamset-class);
-   *msData* refers to the S4 subclass of *streamSet* when MS analyses (i.e., MS files) are added to the set, see \@ref(msdata-class);
-   *msAnalysis* refers to the S4 class of the object generated for each MS file added to the *msData* object in the slot "analyses";
-   `setInfo` is the method to obtain the basic information (i.e., title, date and path) of a *streamSet* object;
-   `analyses` is the name of the slot holding the analyses (as list) and is also a method to get the analysis names in the *streamSet*;
-   `replicate` is the name of an analysis replicate group and the method to get the replicate names;
-   `blank` is the name of a blank analysis replicate group to be used for blank (blind) subtraction and the method to get blank replicate names;
-   **traces** corresponds to a raw data signal (e.g., mass trace in MS data) obtained from a given analysis file;
-   `spectra` for MS analyses is the method to get a table with a spectrum for each cycle time (i.e., retention time);
-   **EIC** is an extracted ion chromatogram (retention time vs intensity) and can be obtained with the method `EICs`;
-   **TIC** is a total ion chromatograms (retention time vs intensity) and can be obtained with the method `TICs`;
-   **XIC** is a three dimensional ion chromatogram (*m/z* vs retention time vs intensity) and can be obtained with the method `XICs`;
-   **peak** corresponds to a cluster of traces (e.g., integrated chromatographic MS peak) within an analysis file and can be accessed with the method `peaks`;
-   **feature** corresponds to a group of corresponding peaks across analysis files/analyses and can be accessed with the method `features`;
-   **MS2** corresponds to second order fragmentation data as acquired in tandem MS or MS/MS mode and can be accessed with the method `MS2s`.

@@

The set creation is demonstrated below for the three MS files (triplicate, 1 to 3 in \@ref(tab:description-files)) where a set of chemicals were spiked (see \@ref(spiked-chemicals) with tag IS and S). The set path can be assigned as argument but for the example below it was left as default (i.e., the working directory). The initial S4 class created in the `newStreamSet` function is *streamSet*. The subclass is then defined according to the type of files added. Note that multiple file types are not possible as a subclass cannot be defined.

@@

Because the added files were all with mzML format (i.e., MS files), the subclass is directly defined as *msData*. The structure for further processing data is then set for MS data. Other data formats/structure will lead to different sub classes (e.g., *uvData* and *ramanData*) but is not yet implemented within the streamFind package.

@@

The *msData* structure includes the *streamSet* class plus the `features` slot. The title, date and path slots correspond to the set title, creation date and directory in disk, respectively. The analyses slot is a list with a *msAnalysis* S4 class object per MS file added. The structure of the *msAnalysis* class is described in the next sub-chapter. The `features` slot is a *msFeatures* S4 class object to hold the results from alignment, grouping and annotation of peaks across analyses. Further information for *msAnalysis* is given in the next sub-chapter (\@ref(msanalysis-class)) and for *msFeatures* is given in the sub-chapter \@ref(msfeatures-class), after creation of features by performing alignment and grouping of peaks across analyses.

-   *msData* structure:

@@

The *msAnalysis* class is structured as shown below. Note that the S4 method `getAnalyses` for *msData* was used to obtain the *msAnalysis* of the first analysis as defined by the second argument. In general, access to raw data (i.e., traces and chromatograms) as well as respective plotting S4 methods can be applied to both the *msAnalysis* and *msData* (as described in \@ref(accessing-raw-data)).

The slot analysis corresponds to the analysis name. The metadata slot is a list of objects with informative entries of the analysis file. The metadata list can be expanded with additional information about the analysis/sample (e.g., sample location, weather conditions, process sensor data, etc.). The slot parameters is used to store processing steps (e.g., peak picking) as *settings* S4 class objects, including the processing step name/ID, the algorithm used and the respective settings applied. The parameters can be used for history track but also for re-run the processing steps when required. The spectra and chromatograms slots are used to store low level data (i.e., mass traces either in profile or centroided modes) from the MS files. Finally, the peaks slot is used to store the chromatographic peaks after peak picking. As no other processing step was applied the spectra, chromatograms and peaks tables in the *msAnalysis* are empty for now.

-   *msAnalysis* structure:

@@

An *msAnalysis* object can be created without the need of running `newStreamSet`. The method `newAnalysis` can be used for initiating the *msAnalysis* when a full path of an MS file (i.e., with extension mzML or other MS format) is given. Similar to `newStreamSet`, the format of the file given as file argument dictates the class of the analysis object returned from the `newAnalysis` function.

@@

As the method `getAnalysis` can be used to obtain an *msAnalysis* from an *msData* object, the method `addAnalyses` is used to add extra *msAnalysis* objects to *msData*. Below the files 7 to 9 are added to the already created *msData* (containing files 1 to 3). Note that both `getAnalyses` and `addAnalyses` can be applied to equivalent classes to *msData* (i.e., *uvData* and *ramanData*). Also, note that when adding analyses to an *msData* object with features (i.e., with peaks grouped across analyses), the *msFeatures* content is cleared as the grouping is invalidated with new *msAnalysis* objects.

@@

Below, the `getMetadata` S4 method of *msData* is used to display the metadata. Note, `getMetadata` also works for *msAnalysis* S4 class as shown below.

-   get metadata:

@@

Additional metadata can be added with the `addMetadata` method for both *msAnalysis* and *msData*. Note that for *msData* the metadata input needs to be a list of vectors with the same length as the number of analyses or a data.frame/data.table with the number of rows as the nmumber of analyses, as shown below.

-   add metadata to an *msAnalysis* object (a1):

@@

-   add metadata to an *msData* object (sp1):

@@

The tools for raw data access and plotting (i.e., spectra and chromatograms) share a target search approach. The data search is based on pre-defined time and mass (inc. *m/z*) targets. For traces (i.e., raw MS mass traces), time (in seconds) and *m/z* (in Da) ranges are given for accessing and plotting. Access to mass traces is relevant for further development of data processing steps/functions and for visualization of raw data, easing the evaluation of processing steps (e.g., peaks wrongly not found during peak picking).

@@

As aforementioned, access to raw data is based on defining *m/z* and retention time targets. The function `makeTargets` is used for comprehensively build targets. As shown below for Carbamazepin d10 and Diuron d6, multiple ways can be used to assemble targets but similarly returning a uniform `data.table` class object with a target list. Accessing and plotting S4 methods in streamFind use the `makeTargets` function to collect raw data from the analysis files. Therefore, the arguments `mz`, `rt`, `ppm`, `sec` and `id` are present in most S4 methods for accessing and plotting data as demonstrated in the following sections.

```{r carb-diu}
#Carbamazepin-d10 is ionized only in positive mode
carbamazepin_d10 <- db[name %in% "Carbamazepin-d10", .(name, mass, rt)]
carbamazepin_d10

#Diuron-d6 is ionized in both positive and negative modes
diuron_d6 <- db[name %in% "Diuron-d6", .(name, mass, rt)]
diuron_d6
```

```{r make_targets}
carb_pos <- carbamazepin_d10$mass + 1.0073
carb_rt <- carbamazepin_d10$rt
diu_pos <- diuron_d6$mass + 1.0073
diu_rt <- diuron_d6$rt

sec_dev <- 30 #seconds
ppm_dev <- 20

#case 1
mz1 <- c(carb_pos, diu_pos)
rt1 <- c(carb_rt, diu_rt)
id1 <- c("target1", "target2")
targets1 <- makeTargets(mz = mz1, rt = rt1, ppm = ppm_dev, sec = sec_dev, id = id1)

@@

#case 2
ppm_carb <- ppm_dev / 1E6 * carb_pos
ppm_diu <- ppm_dev / 1E6 * diu_pos

mz2 <- data.frame(mzmin = c(carb_pos - ppm_carb, diu_pos - ppm_diu),
                  mzmax = c(carb_pos + ppm_carb, diu_pos + ppm_diu))

rt2 <- data.frame(rtmin = c(carb_rt - sec_dev, diu_rt - sec_dev),
                  rtmax = c(carb_rt + sec_dev, diu_rt + sec_dev))

targets2 <- makeTargets(mz = mz2, rt = rt2)
targets2

#case 3
mz3 <- data.frame(
  id = c("target1", "target2"),
  mz = c(carb_pos, diu_pos),
  rt = c(carb_rt, diu_rt)
)

targets3 <- makeTargets(mz = mz3, ppm = ppm_dev, sec = sec_dev)
targets3

#case 4
mz4 <- data.frame(
  id = c("target1", "target2"),
  mzmin = c(carb_pos - ppm_carb, diu_pos - ppm_diu),
  mzmax = c(carb_pos + ppm_carb, diu_pos + ppm_diu),
  rtmin = c(carb_rt - sec_dev, diu_rt - sec_dev),
  rtmax = c(carb_rt + sec_dev, diu_rt + sec_dev)
)

targets4 <- makeTargets(mz = mz4)
targets4
```

@@

Extracted ion chromatograms (EICs) are obtained either from loaded spectra in the *msData* or *msAnalysis* objects or from querying traces directly in the raw data file (mzML or mzXML). The latter is less disk memory intense as data is loaded from raw files on demand. The S4 method `EICs` is used to extract MS1 data from raw data files.

@@

The obtained EICs can be plotted with `plotEICs` S4 method, as shown below. The object for `plotEICs` can be a `data.table` as resulting from the `EICs` method or the actual *msData* or *msAnalysis* objects, using the same arguments as in `EICs` for collecting the data (i.e., to define the targets).

@@

Total ion chromatograms (TICs) are obtained either by pre-loading and adding the chromatogram data to the *msAnalysis*, extracting the data in a `TICs` call (S4 method) or loading all spectra and applying the sum of intensities to each spectrum. Note, TIC is already available as chromatogram in mzML data from the major HRMS vendors. The S4 method TICs first evaluates if the TIC chromatogram is available in the mzML and when not, the spectra data is extracted and used instead.

@@

A more informative method of obtaining EICs is via the method `XICs` which extract a the three dimensional (*m/Z*, rt and intensity) chromatograms from given targets. In the example below, the *m/z* dimension is also included in the resulting table to improve inspection of traces.

@@

Similarly, the XICs can be plotted with the S4 method `plotXICs`, as shown below. Other arguments are available for plotting `XICs`, including a target area for the expected targets. As shown in the plot below, the red mark is defined by the arguments `targetsMark` (a table with columns mz and rt to define the expected *m/z* and retention times) and `ppmMark` and `secMark` to define the mass (in ppm) and time (in seconds) range of the target squares. The `numberRows` is used to control the number of rows of the grouped plot. The concept for the XIC plot is adapted from the R package [MSnbase](https://bioconductor.org/packages/release/bioc/html/MSnbase.html).

@@

The S4 methods applied above parse the data from the raw data files which is less disk memory intense but slower. Alternatively, the raw data can be loaded to (each) *msAnalysis* and from their is faster available. However, when loaded, the raw data has a higher memory footprint. Therefore, the data can be optionally loaded and stored as a simple *data.table* in the respective *msAnalysis* (other classes of the same level might have the raw data directly loaded if the data has a lower memory footprint). Below we demonstrate how to load (via `loadRawData` S4 method) and access (via `spectra` or `chromatograms`) the raw spectra and chromatograms from an *msAnalaysis* and *msData* object. Both S4 methods (`spectra` or `chromatograms`) return a *data.table* where further operations can be applied to subset the data, as demonstrated below. The standard *data.table* operations can be applied, see documentation in <https://cran.r-project.org/web/packages/data.table/> for more information and code examples/instructions. A cheat sheet for *data.table* operations can be downloaded in <https://raw.githubusercontent.com/rstudio/cheatsheets/main/datatable.pdf>. In streamFind, the *data.table* is always the preferred format for table objects.

```{r loadrawdata-msanalysis}
#load raw data to an msAnalysis object, while setting a minimum intensity value for both MS1 and MS2 traces
a1 <- loadRawData(a1, minIntensityMS1 = 0, minIntensityMS2 = 100)
```

@@

The analysis set creation is demonstrated below for twelve HRMS data files corresponding to blank and wastewater influent samples, both measured in triplicate and in positive and negative modes. The argument `files` is the main input and can either be a vector of full path files, a table with file (i.e., full file path), replicate (i.e., the name of the analysis replicate group for each file) and blank (i.e., the name of the associated blank analysis replicate group for each file), or the `analysisInfo` data.frame from the package [patRoon](https://github.com/rickhelmus/patRoon). The set path can be assign as argument `path` and is the directory where all the set files (i.e., scripts, cache databases, objects and results) will be stored by default.

```{r newstreamset-basicworkflow, message=FALSE, results='hide'}
sp2 <- newStreamSet(files = files[10:21], path = getwd(), title = "Project Example")
```

@@

An essential aspect of environmental analysis is to operate with sample/analysis replicates. The replicates in *msData* can be obtained by the `replicateNames` S4 method as shown below. The setter (assignment) for the replicate names is demonstrated below with the method `replicateNames<-` that takes a vector of character strings with the same length as the number of analyses in a *msData* object. Alternatively, a table with a replicate character column or a character vector can be given in the files or replicates arguments, respectively. Note that *msAnalysis* doesn't contain information about replicate correspondence. The correspondence between analyses is storage in the *msFeatures* within the *msData*.

@@

#setter for replicate names in msData
replicateNames(sp2) <- c(
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("influent_neg", 3),
  rep("influent_pos", 3)
)

replicateNames(sp2)
```

@@

#setter for the blank analysis replicate of msData object
blankReplicateNames(sp2) <- c(
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("blank_neg", 3),
  rep("blank_pos", 3)
)

blankReplicateNames(sp2)
```

@@

@@

#getter for polarity of analyses in a msData
polarities(sp2)

@@

`# TODO create function for collecting MS2 data for peaks/features`
